Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-10-01 draft.tex
Final Revision

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
% TODO: Auf 200 Wörter runterkürzen
\abstract{ 
%The goal of epidemiology is to provide insight into disease causations.
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
%We assess the predictive quality of image-derived variables for lower back pain.
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%We assess the predictive quality of image-derived variables for lower back pain.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
%We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g. pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors within the data set.
%%
A correlation between lumbar spine shape and lower back pain was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, associations between non-image variables were successfully detected and characterized.
}

@@

We present an analysis of the image-derived data of a cohort study including lumbar spine datasets.
%%
We extract associations between spine shape and back pain characteristics.
%We analyze their discriminative power for different back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential as back pain risk factors.
Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
%%
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification of lumbar spine shape properties potentially associated with back pain,
%\item The identification and evaluation of lumbar spine shape properties associated with back pain,
%%
\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,
%%
%\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework.
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
%%
For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{risk factor} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
The collected data can be analyzed regarding many diseases and conditions.
%%

@@

Modern cohort studies often comprise medical image data.
%%
The analysis of these data is challenging.
%These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent complaints in the Western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, physical activity and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%
%Potential \emph{confounding} effects are also subject of current research.
%%

@@

\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.

@@

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
%%
The data set comprises 134 variables (and 9 additional image-derived variables described in the next subsection):
\begin{itemize}
	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
	\item 113 categorical variables divided into
	\begin{itemize}
		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. pancreatitis or high blood pressure
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
	\end{itemize}
\end{itemize}
%%
Many variables comprise missing data.
%%
Some variables are follow-up questions, covering the treatment of a disease.
%%
Others are exclusive to subgroups, such as women-specific variables, e.g. menstrual status.
%%
Sparse variables are statistically less resilient and need to be treated with care during the analysis.

\subsection{Image-Derived Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to an MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal $T_1$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

@@

%%
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{I}$. We refer to total curvature as \emph{curvature}.
%%
	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{I}$. From now on, we refer to total torsion as \emph{torsion}.
%%
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This yields a total of 9 image-derived variables.
%%
In the next section, we present the experiments we conducted to assess the influence of the lumbar spine shape to lower back pain.

@@

	Correlation of age and gender regarding the lumbar spine size.

@@

	The bar chart shows subjects divided into four different age groups (x-axis) and their subject number (y-axis).

@@

Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also risk factors (Fig.~\ref{fig:age-gender}).

@@

Large body weight increases the spine load, resulting in a bent shape.

@@

To assess their influence, we take them into account when spine curvature and torsion is correlated with non-image variables.

@@

Since the gender encodes body size, we decided to divide subjects into body size groups.

@@

We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.

@@

A generalized pairs plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.

@@

Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as generalized pairs plot.

@@

The box plots show no difference between subjects with and without back pain.

\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
%%
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
%%
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\paragraph{Heterogenous Correlations.}
%%
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combinations:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
These variables have to be omitted in the analysis, since they are not statistically resilient.
%%
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation of \emph{torsion} with \emph{Parkinson's disease} (0.24).
%%
Other than that, \emph{torsion} correlated with almost no variables (p values between 0 and 0.05).
%%

@@

We apply a square root scale for the tree size axis to highlight data points with few decision rules.
%%
Outlier results with very large decision trees would otherwise distort the resulting plot.
%%
\subsubsection{Dummy Variables}
Dummy variables convert a categorical variable with multiple manifestations into several dichotomous variables.
%%
Each dichotomous variable encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variable.
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on an entry on the visualization, which then interactively displays the corresponding decision tree in detail.
%%
This allows to sequentially analyze the classifications.

@@

\subsection{All Variables}
%%
The vast majority of non-image variables cannot be automatically classified based on the image-variables.
%%
This is reflected in the large amount of variables classified with an error above $0.6$.
%%

None of the pain indicators can be described reliably using the image-based variables.
%%
The only variable reliably classified in this group is gender. % with an error rate of 31.5\% using a decision tree of seven rules.
%
It can be discriminated with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
%%
The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
%%
\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on \emph{coronal mean curvature}.
%%
Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
%%
The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
%%
The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.

@@

The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
%%
\subsection{Size Groups}
%%
Many previously described results are influenced by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is whether the inter-group spine variability parameter is sufficient for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
\emph{Back pain}-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
%%
%TODO: Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
%%
\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
%%
They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
%%
\emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.

>>>>

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g. pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, associations between non-image variables were successfully detected and characterized.
}

@@

We present an analysis of the image-derived data of a cohort study including lumbar spine datasets.
We extract associations between spine shape and back pain characteristics.
%We analyze their discriminative power for different back pain characteristics.
For this purpose, we combine data mining algorithms with data visualization techniques.
%Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential as back pain risk factors.
Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
We focus on highlighting new unexpected correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
\item The identification of lumbar spine shape properties potentially associated with back pain,
%\item The identification and evaluation of lumbar spine shape properties associated with back pain,
\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,
%\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework.
\end{itemize}
The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
\begin{itemize}
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
\end{itemize}
Mutually dependent variables make this analysis challenging.
Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
Age acts as a \emph{risk factor} for prostate cancer.
%These \emph{confounding} variables are often hard to find.
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
The study type depends on the condition of interest.
Most common are case-control studies analyzing one specific disease and its influence on the human body.
We focus on large scale cohort study data.
These studies aim at collecting as much data as possible for each subject.
The collected data can be analyzed regarding many diseases and conditions.

@@

Modern cohort studies often comprise medical image data.
%%
The analysis of these data is challenging.
%These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent complaints in the Western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%
%Potential \emph{confounding} effects are also subject of current research.
%%

@@

\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in tumorous tissue.

@@

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
%%
The data set comprises 134 variables (and 9 additional image-derived variables described in the next subsection):
\begin{itemize}
	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
	\item 113 categorical variables divided into
	\begin{itemize}
		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
	\end{itemize}
\end{itemize}
%%
Many variables comprise missing data.
%%
Some variables are follow-up questions, covering the treatment of a disease.
%%
Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
%%
Sparse variables are statistically less resilient and need to be treated with care during the analysis.

\subsection{Image-Derived Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to an MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

@@

%%
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{I}$. We refer to total curvature as \emph{curvature}.
%%
	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{I}$. We refer to total torsion as \emph{torsion}.
%%
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This yields a total of 9 image-derived variables.
%%
In the next section, we present the experiments we conducted to assess the influence of the lumbar spine shape to lower back pain.

@@

	Correlation of \emph{age} and \emph{gender} regarding the lumbar spine size.

@@

	The bar chart shows subjects divided into four different \emph{age} groups (x-axis) and their subject count (y-axis).

@@

Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors (Fig.~\ref{fig:age-gender}).

@@

Large \emph{body weight} increases the spine load, resulting in a bent shape.

@@

To assess their influence, we take them into account when spine \emph{curvature} and \emph{torsion} is correlated with non-image variables.

@@

Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.

@@

We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes.

@@

A Generalized Pairs Plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.

@@

Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.

@@

The box plots show no difference between subjects with or without back pain.

\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
%%
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
%%
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
%%
The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{transverse curvature}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\paragraph{Heterogenous Correlations.}
%%
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combinations:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
These variables have to be omitted in the analysis, since they are not statistically resilient.
%%
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation of \emph{torsion} with \emph{Parkinson's disease} (0.24).
%%
Other than that, \emph{torsion} correlated with almost no variables (p values between 0 and 0.05).
%%

@@

We apply a square root scale for the tree size axis to highlight data points with few decision rules.
%%
Outlier results with very large decision trees would otherwise distort the resulting plot.
%%
\subsubsection{Dummy Variables}
Dummy variables convert a categorical variable with multiple manifestations into several dichotomous variables.
%%
Each dichotomous variable encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on an entry on the visualization, which then interactively displays the corresponding decision tree in detail.
%%
This allows to sequentially analyze the classifications.

@@

The shown mutual dependencies aim to amplify \emph{hypothesis generation}.
Dedicated statistical analysis of these results solely focussed on epidemiological research is not the scope of this paper.
\subsection{All Variables}
The vast majority of non-image variables cannot be automatically classified based on the image-variables.
This is reflected in the large amount of variables classified with an error above $0.6$.

None of the pain indicators can be described reliably using the image-based variables.
%%
The only variable reliably classified in this group is \emph{gender}. % with an error rate of 31.5\% using a decision tree of seven rules.
%
It can be discriminated with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
%%
The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
%%
\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on \emph{coronal mean curvature}.
%%
Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
%%
The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
%%
The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.

@@

The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
%%
\subsection{Size Groups}
%%
Many previously described results are influenced by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is whether the inter-group spine variability parameter is sufficient for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
%%
%TODO: Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
%%
\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
%%
They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
%%
\emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
